---
title: "Prediction Assignment"
date: "March 29, 2017"
output:
  html_document:
    keep_md: yes
---


## Synopsis

One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, the goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants.

The goal of the project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. We may use any of the other variables to predict with.   

The report describes  

- how we built our model, 
- how we used cross validation, 
- what we think the expected out of sample error is, and 
- why we made the choices we did. 


## Background 
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements abo  ut themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: [The data information][3] (see the section on the Weight Lifting Exercise Dataset).



## Data
The training data for this project are available here:
[The Training Data][1].
  
The test data are available here:
[The Testing Data][2]

The data for this project come from this source [The data information][3]

[1]:https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv   "The training data"
[2]:https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv  "The test data"  
[3]:http://groupware.les.inf.puc-rio.br/har "The data information"


## Summary of Analysis
1. Data exploration (structure, consistency of the column names in both training and testing data)
2. Remove variables of near zero variability
3. Remove variables with more than 50% missing values
4. Remove variables not suitable for analysis, eg. user_name, timestamps, etc.
5. Examine existence of highly correlated pairs
6. 5-fold cross-validation and PCA using trainControl()
7. Build model with multiple methods
8. Compare model performance and select the best model
9. The expected out of sample error
10. Predict on testing data using the best model  


## Analysis

#### Download and import the data into the workspace
```{r download_read_data, cache=TRUE}
trainURL="https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
if(!file.exists("./pml-training.csv")){
  download.file(trainURL, destfile = "./pml-training.csv")}
training<- read.csv("./pml-training.csv",header = TRUE)
testURL="https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
if(!file.exists("./pml-testing.csv")){
  download.file(testURL, destfile = "./pml-testing.csv")}
testing<- read.csv("./pml-testing.csv",header = TRUE)
```


#### Data Exploration 

The data structure of training data and coherence of column names from training and test data
```{r structure_of_data} 
str(training)
#str(testing)
# Are the column names(except classe) consistent between training and testing data
all.equal(colnames(training)[1:length(colnames(training))-1],colnames(testing)[1:length(colnames(testing))-1])
```


#### Feature Extraction
Loading the required libraries,
```{r load_libraries, echo=FALSE}
library(lattice)
library(ggplot2)
library(caret)
library(rattle)
library(dplyr)
library(rpart)
library(corrplot)
```

Keeping just the fields with some variability in the analysis,
``` {r filter_columns}
nearzero<-nearZeroVar(training, saveMetrics = TRUE)
# column names with near zero variability
colnames(training[which(nearzero$nzv)])
# remove columns with near zero variability
train<-training[,!nearzero$nzv]

# column with more than 50% missing values removed
to_remove<-sapply(colnames(train), function(x) if(sum(is.na(train[,x]))>.5*nrow(train)) {return(TRUE)}else{return(FALSE)})
train<-train[,!to_remove]
```
  
Now, remove variables not suitable for analysis: X, user_name,raw_timestamp_part_1, raw_timestamp_part_2, cvtd_timestamp, num_window 
```{r}
train<-train[,-(1:6)]
```
  
The variables to be used for analysis and the summary of the variables
```{r train data}
names(train)
# summary of each variables
summary(train)
# columns with high in pair-wise correlation
hi_corr_columns<-findCorrelation(cor(train[,-53]))
length(hi_corr_columns)
corrplot.mixed(cor(train[,-53]),lower="circle", upper="color", 
               tl.pos="lt", diag="n", order="hclust", hclust.method="complete")
```

#### Cross-validation and PCA
```{r cross validation}
# 5-fold cross validation with trainControl()
trainCtl<-trainControl(method="cv",number=5, preProcOptions="pca",allowParallel=TRUE)
```

#### Models 
Support Vector Machine Radial
```{r svmRadial}
model_svmR<-train(classe~.,method="svmRadial",data=train, trControl=trainCtl)
model_svmR
```
Random Forest
```{r RandomForest}
model_rf<-train(classe~.,method="rf",data=train, trControl=trainCtl)
model_rf
```
RPart
```{r rpart}
model_rpart<-train(classe~.,method="rpart",data=train, trControl=trainCtl)
fancyRpartPlot(model_rpart$finalModel)
model_rpart
```
  
  
#### Model comparison
```{r compare_models}
model=c("rpart","randomForest","svmRadical")
accuracy<-c(max(model_rpart$results$Accuracy),
            max(model_rf$results$Accuracy),
            max(model_svmR$results$Accuracy))
kappa<-c(max(model_rpart$results$Kappa),
         max(model_rf$results$Kappa),
         max(model_svmR$results$Kappa))
modelComparison<-cbind(model,accuracy,kappa)
modelComparison
#knitr::kable(modelComparison)
```
  

#### Model selection
According to the result of each models, the random forest model has the highest accuracy. Now we predict with the random forest model on the testing data.
The top 25 important predictors of the this model is;
```{r model_rf}
plot(varImp(model_rf),xlab="Top 25 Important Features",top=25)
trellis.par.set(caretTheme())
plot(model_rf)
```
  
##### The expected out of sample error
In random forests, there is no need for cross-validation or a separate test set to get an unbiased estimate of the test set error. It is estimated internally, during the run, as follows:

Each tree is constructed using a different bootstrap sample from the original data. About one-third of the cases are left out of the bootstrap sample and not used in the construction of the kth tree.

Put each case left out in the construction of the kth tree down the kth tree to get a classification. In this way, a test set classification is obtained for each case in about one-third of the trees. At the end of the run, take j to be the class that got most of the votes every time case n was oob. The proportion of times that j is not equal to the true class of n averaged over all cases is the oob error estimate. This has proven to be unbiased in many tests.[Reference][4]  

## Prediction
``` {r prediction on testing data}
pred<-predict(model_rf,testing)
pred
```


[4]:https://www.stat.berkeley.edu/~breiman/RandomForests/cc_home.htm "Reference"
